---
title: "Tracking Atlantic Bluefin Tuna Across EEZ Boundaries"
editor: visual
---

## Background

### Atlantic Bluefin Tuna

**Atlantic Bluefin Tuna** (*Thunnus thynnus*) are among the largest and fastest fish in the ocean - adults can exceed 3 meters in length and 600 kg, and sustain swimming speeds over 70 km/h. They are highly migratory, crossing the Atlantic Ocean multiple times throughout their lives. Two spawning populations are recognized: one that spawns in the Gulf of Mexico and another in the Mediterranean Sea. Outside the spawning season, individuals roam widely across the North Atlantic in search of prey, making them one of the most wide-ranging fish species known.

Bluefin tuna are also one of the most commercially valuable fish in the world. A single large individual can sell for tens of thousands of dollars at auction. This economic value, combined with their broad geographic range, makes managing bluefin populations exceptionally challenging - and understanding *where* they go is fundamental to any management strategy.

### Electronic Tagging

To track bluefin tuna movements, scientists attach **electronic tags** that record depth, temperature, and light levels. Light-level data allow researchers to estimate the animal's longitude (from the timing of dawn and dusk) and latitude (from day length). These raw estimates are noisy - errors of hundreds of kilometers are common - so scientists use statistical models to filter the raw observations into a smoothed daily track. The model accounts for measurement error and the fact that an animal's true movement must be physically plausible (it cannot teleport). The tracks you will work with today are statistically-estimated daily positions for 19 Atlantic Bluefin Tuna tagged in the North Atlantic.

### Exclusive Economic Zones

Coastal nations claim an **Exclusive Economic Zone (EEZ)** extending 200 nautical miles (≈ 370 km) from their shorelines. Within this zone, the nation has sovereign rights over living and non-living resources, including fisheries. Because bluefin tuna migrate freely across EEZ boundaries, effective conservation requires coordination among many nations. Knowing *how much* of a tuna's migration falls within each country's EEZ is key information for negotiating quotas and assigning management responsibility.

------------------------------------------------------------------------

## Setup

Load `tidyverse`, `terra`, and `tidyterra`. Then read in the three spatial datasets:

-   `data/tracks/tracks.shp` - tuna tracks as lines, one geometry per individual
-   `data/eez/eez.shp` - EEZ polygons
-   `data/land/land.shp` - country land polygons for basemap reference

```{r}
library(tidyverse)
library(terra)
library(tidyterra)

# Replace the NULLs below.
tracks <- NULL
eez    <- NULL
land   <- NULL
```

------------------------------------------------------------------------

## Part 1: Map the tracks

### Explore the data

**Q1.** Use `nrow()` on `tracks` to find out how many tuna are in the dataset. Then use `names()` to see what attributes are attached to each track geometry.

```{r}
# Your code here
```

*Answer:*

**Q2.** What attribute identifies each individual tuna?

*Answer:*

### Build the map

Create a map showing land, EEZ boundaries, and tuna tracks. You will need:

-   `geom_spatvector()` - plots a `SpatVector` layer; use it for all your data sources (no rasters).
-   `coord_sf()` - sets the map extent; use `xlim` and `ylim` to zoom to the data.
-   Color each track by `TOPPID`.

```{r}
# Your code here
```

**Q3.** Visually, which parts of the North Atlantic do the tracks concentrate in? Do any tracks appear to stay within a single EEZ, or do most cross boundaries?

*Answer:*

------------------------------------------------------------------------

## Part 2: Time spent in each EEZ

To quantify how much of each migration falls inside each EEZ, we will use `intersect()` to clip the track lines to the EEZ polygons, then measure the length of each clipped segment with `perim()`.

### Clip tracks to EEZs

`intersect()` returns a new `SpatVector` containing only the portions of each track that fall inside an EEZ polygon. The result inherits attributes from both inputs - each row will have a `TOPPID` (from `tracks`) and a `GEONAME` (from `eez`).

After intersecting, use `perim()` to compute the length of each clipped segment. `perim()` takes a `SpatVector` as its argument and returns a vector of lengths - one value per geometry row. Create a new column called `length` with the length of each segment.

```{r}
# Your code here
in_eez <- NULL   # tibble with columns TOPPID, GEONAME, length
```

**Q4.** How many rows does `in_eez` have? What does each row represent?

```{r}
# Your code here
```

*Answer:*

### Compute fractions

Within each `TOPPID` group, compute `fraction` as the fraction of each segment's length relative to the total length in the group. This is the share of EEZ-tracked migration spent in each EEZ.

```{r}
# Your code here
zone_fractions <- NULL
```

**Q5.** Which tuna visits the most distinct EEZs? Fill in the table below.

```{r}
# Your code here
```

| TOPPID | EEZs visited |
|--------|--------------|
|        |              |

------------------------------------------------------------------------

## Part 3: Per-EEZ total migration

Now summarize across all tunas: for each EEZ, sum `length` across individuals to get total track length. Sort in descending order and filter out EEZs that represent less than 5% of the total (these are too small to show clearly).

Make a bar chart with EEZ on the x-axis and total track length on the y-axis.

```{r}
# Your code here
eez_totals <- NULL
```

```{r}
# Your code here
```

**Q6.** Which EEZ accounts for the most total tuna migration across all individuals?

*Answer:*

------------------------------------------------------------------------

## Submission

You've completed this week's lab. Follow these instructions to submit to Gradescope.

1.  **Render this Quarto document.** Verify that it renders without errors and that your figures appear in the output HTML.
2.  **Download tunatracks.qmd to your computer.**
3.  **Navigate to the Gradescope assignment.** There's a link in the Canvas assignment.
4.  **Click** [**Submit Assignment**]{.underline}**, then drag and drop your Quarto file** (or click to browse for it).
